---
title: "R Gym Trainer"
author: "Tanveer Ansari"
date: "Saturday, May 23, 2015"
output: html_document
---
### Who needs a personal trainer when you have R 
###  and the IOT(Internet of Things)

This project attempts to predict if test subjects are performing barbell curls correctly   
and if not to predict what type of mistake they are doing. This information is taken  
from [Human Activity Recognition](http://groupware.les.inf.puc-rio.br/har

```{r IMPORTS,results='hide',message=FALSE,warning=FALSE,include=FALSE}
require(dplyr); require(data.table);require(RCurl)
require(caret);require(rattle);require(knitr);require(rpart);
```

`r opts_chunk$set(cache=TRUE)`

### Load and Clean Data
```{r DOWNLOAD_DATA, cache=TRUE}
if(!file.exists("train.rds")){
  trainURL<-getURL("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
  trainAll<-read.csv(text=trainURL)
  saveRDS(trainAll,"train.rds")
  } else {
    trainAll<-readRDS("train.rds") ##Training dataset
    }

if(!file.exists("test.rds")){
  testURL<-getURL("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
  testingFinal<-read.csv(text=testURL)
  saveRDS(testingFinal,"test.rds")
  } else {
    testingFinal<-readRDS("test.rds")  ## Testing dataset
    }
```


**Remove irrelevant columns  **  
1. Remove the column X - this is the order in which the data is presented  
2. Remove the user_name and num_window columns - I do not believe they are relevants as factors  
3. Remove all timestamp columns - this analysis did not treat this data as time series data  

```{r REMOVE_USELESS_VARS}
tidyTrain<-trainAll
tidyTrain$X<-NULL # remove sequence number from data
tidyTrain$user_name<-NULL
tidyTrain$num_window<-NULL
# remove timestamp columns from dataset
tidyTrain<-tidyTrain[,-grep("time",names(tidyTrain))]
```

**Remove fields with minimal variance** - most likely these are not good predictors
```{r REMOVE_NEAR_ZERo_VARIANCE}
nzv<-nearZeroVar(tidyTrain)
removedCols<-names(tidyTrain)[nzv] # Print names of removed columns
tidyTrain<-(tidyTrain[,-nzv])
head(removedCols)
```

**Remove fields that are mostly NAs.**  
I chose 97% NAs as the threshold for removal
```{r REMOVENULLS}
nullMoreThan97<-apply(tidyTrain,2,function(x){as.logical(sum(is.na(x))/length(x) > 0.975)})
#which(nullMoreThan97)
tidyTrain<-tidyTrain[,-which(nullMoreThan97)]
```

### Split data for cross validation   

The training data was split up into a training and a test set in a 75/25 split
```{r CROSSVALIDATION}
set.seed(34455) # Set seed so results are reproducible
trainIndex <- createDataPartition(y=tidyTrain$classe, p=0.75, list=FALSE)
training<- tidyTrain[trainIndex,]
testing<- tidyTrain[-trainIndex,]
```

### Model Fitting  

#### Decision Tree Models

I first fit a decision tree model to the training subset and evaluated its accuracy
```{r DECISION_TREE}
rpMdl<-train(classe~.,method="rpart",data=training) # Train Model 
testing$predicted <- predict(rpMdl, newdata = testing) # Predict on testing subset
fancyRpartPlot(rpMdl$finalModel)
```

This looks promising, lets check accuracy
```{r DECISION_TREE_ACCURACY}
table(testing$predicted,testing$classe)
confusionMatrix(testing$predicted,testing$classe) # Evaluate prediction accuracy on test set
```

The accuracy was ~49% for my run - terrible.  
Next lets fit another decision tree , this time using rpart2 instead of rpart

```{r DECISION_TREE2}
rpMdl<-train(classe~.,method="rpart2",data=training) # Train Model 
testing$predicted <- predict(rpMdl, newdata = testing) # Predict on testing subset
confusionMatrix(testing$predicted,testing$classe)
```

This model showed an improved accuracy of ~55% - better but still terrible

#### Random Forest Model  

Next git a random forest model. This took several hours to run so I am using the
saved model for this presentation.

```{r RANDOM_FOREST_TRAINING}
if(file.exists("RandomForestModelFit")) {
  rfMdl<-readRDS("RandomForestModelFit")
  } else {
    rfMdl<-train(classe~.,method="rf",data=training) # Train Model 
    }

testing$predicted<-predict(rfMdl, newdata = testing)# Predict on testing subset
table(testing$predicted,testing$classe) # Evaluate prediction accuracy on test set
confusionMatrix(testing$predicted,testing$classe) 
```

The random forest model showed an accuracy of ~ 99.76% with a confidence interval   
of (99.57%, 99.87%).

**Out of Sample Error was (1-0.9976) = 0.0024 or 0.24% **

```{r RANDOM_FOREST_PLOT}
plot(rfMdl, log="y")
#varImpPlot(rfMdl)
```

### FINAL MODEL FOR PREDICTION  
 For the final model - train it on the entire test dataset available, not just the training split

```{r FINAL_MODEL}
if(file.exists("RandomForest")) {
  rfMdlFinal<-readRDS("RandomForest")
} else {
  rfMdlFinal <-train(classe~., method="rf",data=tidyTrain)
}

# Predict testing set with Random Forest Model
testingFinal$classe<-predict(rfMdlFinal,newdata=testingFinal)
```

### CONCLUSION

The random forest model, without any preprocessing or resampling during model fit predicted all 20 items in the test set correctly.
